---
title: "Quiz2"
author: "Yuting"
date: "2020/4/24"
output:
  word_document:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    theme: readable
    toc: yes
params:
  release: no
  solution: yes
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
library(knitr)
opts_chunk$set(echo=FALSE,
               cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE)

grd = FALSE
```

## Topic: Monte-Carlo integration

### Problem description:
Write a function use the Monte-Carlo integration to get the integration on [a, b] for arbitray function. 

$I(f)=\int_{a}^{b}(f(x))$, arbitrary f.
Note: (a,b) is an arbitrary interval, finite or infinite
Assume finite.

### Solution:
Use Monte-Carlo integration approach to estimate the integration, standard error and 95% CI.
```{r, echo=TRUE, eval=TRUE}
#Function to be integrated over [a,b]
f <- function(x){x} 

Integral <- function(n,a,b,h=f){
u <- runif(n,a,b) 
Y <- (h(u))/(1/(b-a))
Int <- mean(Y)

u <- runif(n,a,b)
YY <- (h(u)/(1/(b-a)))^2
SE <- sqrt((mean(YY)-Int^2)/n)
CI <- c(Int-1.96*SE,Int+1.96*SE)
list("Int"=Int,"SE"=SE, "CI"=CI)}

```

### Evaluation:

Apply the function on both uniform distribution and exponential distribution.
```{r, echo=TRUE, eval=TRUE}
Integral(1000,0,1,f)

f1 <- function(x){exp(-x)}
Integral(1000,0,10,f1)
```

### Reference
[Monte Carlo Integration](https://cran.r-project.org/web/packages/SI/vignettes/my-vignette.html)
